---
title: 'STA 445 Assignment #7'
author: "Chip Haskins"
date: "2024-11-06"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(latex2exp)
```

### Exercise 2 {-}

Using the `datasets::trees` data, complete the following. This question refreshes create a linear model, graphing the linear model, and introduces using some LaTeX expressions on the graph.

**a)** Create a regression model for $y=$ `Volume` as a function of $x=$ `Height`.

```{r}
trees.lm <- lm(Volume ~ Height, trees)
trees.lm
```

**b)** Display the `summary` of the  model to view the y-intercept and slope of the regression line.

```{r}
summary(trees.lm)
```

**c)** Using `ggplot2`, create a scatter plot of Volume vs Height. 

```{r}
ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point()
```

**d)** Color the scatter using the `Girth` variable.

```{r}
ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth))
```

**e)** Modify the color scheme using a RColorBrewer palette.

```{r}
ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth)) +
  scale_color_distiller(palette = "Greens")
```

**f)** Create a nice white filled rectangle to add text information. The following might be useful.

```{r}
ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth)) +
  scale_color_distiller(palette = "Greens") +
  annotate('rect', xmin=65, xmax=75, ymin=60, ymax=74,
         fill='white', color='black')
```

**g)** Use the `broom` package to extract the coefficients of the best-fit line. Add this information as an annotation to the graph, which should follow a form that looks like $\hat{y}_i = (INTERCEPT) + (SLOPE)* x_i$. Place the annotation within the white text box.

```{r}
trees.lm.coefs <- broom::tidy(trees.lm)
trees.lm.intercept <- filter(trees.lm.coefs, term == "(Intercept)") %>%
  pull(estimate) %>%
  round(3)
trees.lm.slope <- filter(trees.lm.coefs, term == "Height") %>%
  pull(estimate) %>%
  round(3)

ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth)) +
  scale_color_distiller(palette = "Greens") +
  annotate('rect', xmin=65, xmax=75, ymin=60, ymax=74,
         fill='white', color='black') +
  annotate("text", x = 70, y = 67, 
           label = TeX(r"($\hat{y}_i = \intercept + \slope*x_i$)",
                       user_defined = list("\\intercept" = trees.lm.intercept,
                                           "\\slope" = trees.lm.slope)))
```

**h)** Use the `broom` package to extract the coefficient of determination $r^2$ from the model. Add the annotation to your graph, which should look something like $R^2 = (VALUE)$

```{r}
trees.lm.rsquared = broom::glance(trees.lm) %>%
  pull(r.squared) %>%
  round(3)

ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth)) +
  scale_color_distiller(palette = "Greens") +
  annotate('rect', xmin=65, xmax=75, ymin=60, ymax=74,
         fill='white', color='black') +
  annotate("text", x = 70, y = 69, 
           label = TeX(r"($\hat{y}_i = \intercept + \slope*x_i$)",
                       user_defined = list("\\intercept" = trees.lm.intercept,
                                           "\\slope" = trees.lm.slope))) +
  annotate("text", x = 70, y = 65, 
           label = TeX(r"($R^2 = \rsquared$)",
                       user_defined = list("\\rsquared" = trees.lm.rsquared)))
```

**i)** Add the regression line in red. There are several ways to do this.

```{r}
trees.predicted <- broom::augment(trees.lm, trees)

ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth)) +
  scale_color_distiller(palette = "Greens") +
  annotate('rect', xmin=65, xmax=75, ymin=60, ymax=74,
         fill='white', color='black') +
  annotate("text", x = 70, y = 69, 
           label = TeX(r"($\hat{y}_i = \intercept + \slope*x_i$)",
                       user_defined = list("\\intercept" = trees.lm.intercept,
                                           "\\slope" = trees.lm.slope))) +
  annotate("text", x = 70, y = 65, 
           label = TeX(r"($R^2 = \rsquared$)",
                       user_defined = list("\\rsquared" = trees.lm.rsquared))) +
  geom_line(data = trees.predicted, aes(y = .fitted), color = "red")
```

**j)** Properly label the axes of the graph.

```{r}
ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth)) +
  scale_color_distiller(palette = "Greens") +
  annotate('rect', xmin=65, xmax=75, ymin=60, ymax=74,
         fill='white', color='black') +
  annotate(
    "text", 
    x = 70, 
    y = 69, 
    label = TeX(
      r"($\hat{y}_i = \intercept + \slope*x_i$)",
      user_defined = list("\\intercept" = trees.lm.intercept,
                          "\\slope" = trees.lm.slope)
      )
    ) +
  annotate(
    "text", 
    x = 70, 
    y = 65, 
    label = TeX(
      r"($R^2 = \rsquared$)",
      user_defined = list("\\rsquared" = trees.lm.rsquared)
      )
    ) +
  geom_line(data = trees.predicted, aes(y = .fitted), color = "red") +
  labs(
    x = "Tree height (ft)", 
    y = TeX("Volume of timber produced $(ft^3)$"), 
    color = "Trunk diameter (in)"
    )
```

**k)** Add a descriptive title to the graph.

```{r}
ggplot(trees, aes(x = Height, y = Volume)) +
  geom_point(aes(color = Girth)) +
  scale_color_distiller(palette = "Greens") +
  annotate('rect', xmin=65, xmax=75, ymin=60, ymax=74,
         fill='white', color='black') +
  annotate(
    "text", 
    x = 70, 
    y = 69, 
    label = TeX(
      r"($\hat{y}_i = \intercept + \slope*x_i$)",
      user_defined = list("\\intercept" = trees.lm.intercept,
                          "\\slope" = trees.lm.slope)
      )
    ) +
  annotate(
    "text", 
    x = 70, 
    y = 65, 
    label = TeX(
      r"($R^2 = \rsquared$)",
      user_defined = list("\\rsquared" = trees.lm.rsquared)
      )
    ) +
  geom_line(data = trees.predicted, aes(y = .fitted), color = "red") +
  labs(
    title = "Volume and Diameter of Black Cherry Trees by Height",
    x = "Tree height (ft)", 
    y = TeX("Volume of timber produced $(ft^3)$"), 
    color = "Trunk diameter (in)"
    )
```